---
title: "Cobalt Paper"
date: "Updated on : `r date()`"
output: html_document
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Goal

The goal of this vignette is to replicate the analysis presented in ``Association of blood cobalt concentrations with dyslipidemia, hypertension, and diabetes in a US population
A cross-sectional study; Hongxin Wang, MD, Feng Li, MD, Jianghua Xue, MD, Yanshuang Li, MD, Jiyu Li, MD''

 The authors report using data for the years 2015 to 2018 which cover two of the two-year reporting epochs in NHANES.
The Questionnaires we want will have the suffixes _I and _J. 

## 1. Load libs

```{r setup,warning=FALSE,message=FALSE}
library(splines)
library(ggplot2)
library(ggpubr)
library(dplyr)
#library(nhanesA)
library(phonto)
library(EnWAS)
library(knitr)
```


## 2. Data and Preprocessiing
#### 2.1) Blood Pressure Data

 In the next code segment we show how to obtain blood pressure measurements from NHANES tables.  We have
already ascertained that these measurements are contained in Questionnaires that are named BPX_? and that
there were replicate measurements taken.  Both systolic (BPXS) and diastolic (BPXD) measurements were taken
on each occassion.  We
```{r data_blood,warning=FALSE,message=FALSE}
bptablenames = nhanesSearchTableNames('BPX[_]')
bptablenames |> kable()
```
We can see that blood pressure data was collected for other years as well, but for now we will just extract
the data for the 2015-2016 and 2017-2018 years. We combine these into a single dataframe.

```{r extract_data, warning=FALSE, message=FALSE}
blood_df <- unionQuery(c("BPX_I", "BPX_J"), cols =c("BPXDI1","BPXDI2","BPXSY1","BPXSY2"))
dim(blood_df)
blood_df[1:10,] |> kable()
```
  In our analysis we can then look at the average of the measurements across the two different time points as a way to
better estimate the actual blood pressure for each participant.

#### 2.2) Demo, Body Mesures, and Blood Pressure & Cholesterol 

The authors state: 
Participants with cobalt and lipid data were included (n = 6866). Demographic characteristics of the participants, including age, gender, body mass index (BMI), education level, race, family poverty-income ratio and smoking status, were collected. Clinical data, such as blood pressure, total cholesterol (TC), low-density lipoprotein cholesterol (LDL-C), HDL-C, triglycerides (TGs), hypertension, diabetes and history of medication use, including antihypertensive drugs, hypoglycemic drugs, and lipid-lowering drugs, were extracted.

In the calls below, we start with the variable names, which we had obtained by searching based on the variable descriptions (not shown) and restrict by the years that the authors had chosen.

```{r demo_body,warning=FALSE,message=FALSE}

##get the appropriate table names for the variables we will need
##BP
BPTabs = nhanesSearchVarName("BPQ050A", ystart="2015", ystop="2018")
LDLTabs = nhanesSearchVarName('LBDLDL',ystart="2015", ystop="2018")
##A1C
A1C = nhanesSearchVarName("LBXGH",ystart="2015", ystop="2018")
##been told by Dr. has diabetes
DrDiab = nhanesSearchVarName("DIQ010",ystart="2015", ystop="2018")
##HDLTabs
HDLTabs = nhanesSearchVarName("LBDHDD",ystart="2015", ystop="2018")
BMITabs = nhanesSearchVarName("BMXBMI", ystart="2015", ystop="2018")
BMXTabs = nhanesSearchVarName("BMXBMI",ystart="2015", ystop="2018")
DIQTabs = nhanesSearchVarName("DIQ010",ystart="2015", ystop="2018")
COBTabs = nhanesSearchVarName("LBXBCO",ystart="2015", ystop="2018" )

cols = list(DEMO_I=c("RIDAGEYR","RIAGENDR","RIDRETH1","DMDEDUC2","years"), 
            DEMO_J=c("RIDAGEYR","RIAGENDR","RIDRETH1","DMDEDUC2","years"),
            BPQ_I=c('BPQ050A','BPQ020'),BPQ_J=c('BPQ050A','BPQ020'), 
            HDL_I=c("LBDHDD"),HDL_J=c("LBDHDD"), TRIGLY_I=c("LBXTR","LBDLDL"),
            TRIGLY_J=c("LBXTR","LBDLDL"),GHB_I="LBXGH",GHB_J="LBXGH",DIQ_I="DIQ010",
            DIQ_J="DIQ010", BMX_I="BMXBMI", BMX_J="BMXBMI", CRCO_I="LBXBCO",CRCO_J="LBXBCO"
            )
var2Table = cols[c(1,3,5,7,9,11,13)]
base_df <- jointQuery(cols)

```
### 2.3)  merge and PHESANT-like process
 
We can run the PHESANT-like process to convert each column into data types. It also provides the ratio of unique values, the proportion of zeros, and the ratio of NAs, which is calculated by the number of unique values, zeros, and NAs divided by total records. The categorical data types (ordered or unrecorded) are presented by integers, and the PHESANT function category them as multilevel. For example, education (DMDEDUC2) is category as Multilevel-7 means the PHESANT process considers it multilevel and has 7 levels. 

```{r data,warning=FALSE,message=FALSE}

data <- merge(blood_df,base_df,by="SEQN")
##fix up our list linking variable names to the table they came from
var2Table = c(var2Table, list("BPX_I"=c("BPXDI1","BPXDI2","BPXSY1","BPXSY2")))
phs_dat = phesant(data)
data = phs_dat$data
DT::datatable(phs_dat$phs_res)

```

In the next code chunk we will convert the multilevel variables into R factors. To do this we make use of the nhanesTranslate function. While that function does most of the work, there are some variables we need to deal with manually to replicate the analysis. First, the years variable is something we added, so we need to deal with it manually.  Then we use `nhanesTranslate` to transform all of the internal values.  Then we need to do a little more work on modifying levels of the education variable.  The authors decided to group education into three levels, less than high school (<HS), high school (HS), and more than school-school(>HS).  We also need to address some issues around the hypertension variables.
Particpants were asked (Q010)
- The blood pressure measurements (diastolic and systolic) are averaged from the first and second reads.

```{r}
##years is something we have added so we need to translate it directly
data$years <- as.factor(data$years)
##almost surely not the most efficient way
for(tabname in names(var2Table)) {
  data = nhanesTranslate(tabname, var2Table[[tabname]], data)
}
levsED = levels(data$DMDEDUC2)
##examine them and then put into groupings <HS, HS and > HS
levels(data$DMDEDUC2) <- c("HS",">HS","<HS",">HS","<HS",NA)
##
#the variables BPQ050A and BPQ040A are taken conditionally so they have missing values put in
# for anyone who answered no to BPQ020 - and we need to fix that - since there answer to 50A 
 # had they been asked, would have been no (one presumes)
 #set up a dummy for the ones we want to change - also turn "Don't know" from BPQ020 into NA
  gg = is.na(data$BPQ050A) & data$BPQ020=="No"
  bpq50A = data$BPQ050A
  bpq50A[gg] = 2
 #check it worked
 # table(bpq50A, data$BPQ050A, useNA="always")
  
# The current study included 6866 US adults aged 40 years or older. 
data$BPQ050A = bpq50A
##remove any record with at least one NA and then subset to those over 40
data <- na.omit(data)
data <- data[data$RIDAGEYR>=40,]

# Average the the first and second reads
data$DIASTOLIC <- (data$BPXDI1+data$BPXDI2)/2
data$SYSTOLIC <- (data$BPXSY1+data$BPXSY2)/2
```

## 2.4 Hypertension
We follow the definition they described in the paper, using reported systolic and diastolic blood pressure measurements as well as self-reported statements regarding whether a physician had ever told them that they have high blood pressure.  

Note that it is unclear whether the authors used averaged over 2 measurements for the systolic and diastolic blood pressure measurements. Still, we use average them because it would give us more accurate blood pressure measurements.

One might also look at the use of prescribed hypertensives, as these will modulate the systolic and diastolic measures.  Data on self-report come from the BPQ tables in NHANES.
https://wwwn.cdc.gov/nchs/nhanes/2011-2012/BPQ_G.htm

```{r Hypertension,warning=FALSE,message=FALSE}
# "Hypertension was defined as systolic blood pressure (SBP) ≥140 mm Hg, diastolic blood pressure ≥90mm Hg, or the use of antihypertensive medication. "
data$hypertension <- data$DIASTOLIC >= 90 | data$SYSTOLIC >= 140 |  data$BPQ050A=="Yes"
barplot(table(data$hypertension))
```

## 2.5 Compare with Table-2
```{r, message=FALSE,warning=FALSE}
pcobalt = ifelse(data$LBXBCO <= 0.12, "<=0.12", 
                ifelse(data$LBXBCO >= 0.13 & data$LBXBCO <= 0.14, "0.13-0.14",
                  ifelse(data$LBXBCO >= 0.15 & data$LBXBCO <= 0.18, "0.15-0.18",
                         ifelse(data$LBXBCO >= 0.19, ">=1.9",
                         NA)  )))
data$pcobalt = pcobalt

DBP = data |> group_by(pcobalt) |> summarise(mean=mean(DIASTOLIC),SD=sd(DIASTOLIC))
DBP$stat = paste(round(DBP$mean,1),"±",round(DBP$SD,1))

SBP = data |> group_by(pcobalt) |> summarise(mean=mean(SYSTOLIC),SD=sd(SYSTOLIC))
SBP$stat = paste(round(SBP$mean,1),"±",round(SBP$SD,1))

dbp_t = t(DBP)
colnames(dbp_t) = DBP$pcobalt

sbp_t = t(SBP)
colnames(sbp_t) = SBP$pcobalt

table2 = rbind(sbp_t["stat",],dbp_t["stat",])
table2 = table2[,c("<=0.12","0.13-0.14","0.15-0.18",">=1.9")]
table2 = cbind("Blood Pressures"=c("SBP (mm Hg), mean±SD","DBP (mm Hg), mean±SD"),table2)
```
It shows the number we have is not exactly the same as the one in the table-2 in the paper. The authors did not use the average of two reads of the blood pressure measurements.

```{r, message=FALSE,warning=FALSE}
library(kableExtra)
kbl(table2) |>
  kable_classic() |>
  add_header_above(c(" " = 1, "Cobalt Quartiles (ug/L)" = 4))

```





## 3.Base Models and QA/QC

In the following section, we run the logistic regression models as Generalized Linear Models (GLM). In the models, the outcome of the hypertension indicator and the adjusted variables are age(RIDAGEYR), gender(RIAGENDR), BMI(BMXBMI), education(DMDEDUC2), and ethnicity (RIDRETH1). The first GLM is with linear terms, and the second GLM also adds terms linearly together but applies a natural spline to the continuous variables.      

```{r QA/QC, warnings=FALSE, message=FALSE}
## glm linear in the covariates

lm_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+RIDRETH1, data = data, family = "binomial",na.action=na.omit)

##spline covariates
ns_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7)+RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1, 
                   data = data, family = "binomial",na.action=na.omit)

```
### 3.1) QA/QC
```{r}
# library(pROC)
library(plotROC)
test = data_frame(hypertension=data$hypertension,lm=lm_logit$fitted.values,ns=ns_logit$fitted.values)
longtest <- reshape2::melt(test,id.vars="hypertension")
colnames(longtest) = c('hypertension','model','value')
ggplot(longtest, aes(d = as.numeric(hypertension), m = value, color = model))+ geom_abline()+ geom_roc(size = 1.25) + style_roc()

# plot(roc(data$hypertension,
#                    fitted(lm_logit)),
#                print.auc = T, 
#                col = "red")
# 
# plot(roc(data$hypertension,
#                    fitted(ns_logit)),
#                print.auc = T, 
#                col = "blue", 
#                add = T)
```

```{r}
# Age
df_age_fitt = list("Binned Data"=make_bins(x=data$RIDAGEYR,y=as.numeric(data$hypertension),nbin=600),
                  "Linear"=make_bins(x=data$RIDAGEYR,y=lm_logit$fitted.values,nbin=600),
                  "Spline"=make_bins(x=data$RIDAGEYR,y=ns_logit$fitted.values,nbin=600)
                )
age_fitt = plot_bins2(df_age_fitt,xlab="Age (year)",ylab="Hypertension",is_facet=F) 

#BMI
df_bmi_fit =list("Linear"=make_bins(x=data$BMXBMI,y=lm_logit$fitted.values,nbin=600),
                "Spline"=make_bins(x=data$BMXBMI,y=ns_logit$fitted.values,nbin=600),
                "Binned Data"=make_bins(x=data$BMXBMI,y=as.numeric(data$hypertension),nbin=600)
                )

bmi_fit <- plot_bins2(df_bmi_fit,xlab="BMI",ylab="Hypertension",is_facet=F) 

```
The following plots show binned Hypertension data; each bin contains about 600 data points and we compute the proportion of the participants who reported hypertension. 
Linear and Spline present the fitted values (probabilities) from the GLM with linear terms and apply the natural spline function on continuous terms of the participants who have 
hypertension. For both age, panel a), and BMI, panel b),  the GLM model using splines agrees with the estimates obtained by binning, while when these terms are modeled using
a simple linear term there are more substantial discrepancies. 

To compute the model estimates for each bin we simply average the computed fitted values (which are defined to be back-transformed to probabilities for logistic regression) over the 
same individuals in each bin. One might want to examine the relationship on the logit scale, which is easily done.

```{r qa_qc_plot, echo=TRUE,warning=FALSE,message=FALSE, fig.width = 12,fig.height=6}

ggpubr::ggarrange(age_fitt,bmi_fit,nrow = 1,ncol = 2,labels = c('a)','b)'))
```


## 4. Their findings
As the authors pointed out, the blood cobalt concentrations have not associated with the risk of hypertension based on the following summary table. The cobalt concentration does not significantly impact hypertension.
FIXME: but they have a bunch of other features in their table 2 - and it would be good if we can start to look at them.

```{r model1,warning=FALSE,message=FALSE}
lm_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+RIDRETH1+LBXBCO, data = data, family = "binomial")
ns_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7)+RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1+LBXBCO, 
                   data = data, family = "binomial",na.action=na.omit)

sjPlot::tab_model(lm_logit,ns_logit,
                  dv.labels = c("lm", "spline"),
                  show.ci = FALSE,show.stat = TRUE,show.se = TRUE,p.style = "scientific", digits.p = 2)

```
  FIXME:  This likely belongs in the QA/QC section.  The point of this code is to show the reader how they can estimate the functional form of the spline that they are fitting
to the data.  To do that, we pick a covariate, say Age, where we want to compute the spline.  Then we pick a set of Age values that cover the range of ages in the model.  
To get predictions from the model for a specific age we also need to specify values for all the other covariates in the model.  
Our suggestion is that for categorical variables choose the most common category and for continuous variables use the median value.

  FIXME: below is code from RG - we probably want to be using spline functions for AGE and BMI - and then checking those functions.
  We are missing a bunch of the covariates in the Cobalt paper - specifically things like triglycerides etc from Table 2 in that paper.
Those would be interesting to look at, as they seem to have an unusual pattern of association.

*** Robert, you only keep one base model in the end, do you want to me plot or compare somethings***
FIXME: yes the point here is to develop a better summary of the comparison of models with spline terms. I really don't like the R output that shows each term individually, as you can't really interpret them and they take up a lot of room.  I think we should instead, create one line for each spline term, and in it only put the value of the LRT comparing the model with the spline to the model without it.  From that comparison you can get the chi-squared statistic, the p-value and the df and those could be put into the table.  I think that would be a better thing.


```{r model2,warning=FALSE,message=FALSE}
   # base_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+RIDRETH1 + sqrt(LBXBCO), data = data, family = "binomial")
   # base_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7) + RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1 + ns(sqrt(LBXBCO), df=7) + years, data = data, family = "binomial")
   base_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7) + RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1, data = data, family = "binomial")

   ## try to do some prediction - and then get a plot of the age spline
  ##46 of these
  yvals = seq(40,85,by=1)
  dfimpute = data.frame(RIDAGEYR=yvals, RIAGENDR=rep("Male", 46), BMXBMI=rep(28.9, 46), DMDEDUC2=rep("HS", 46), RIDRETH1=rep("Non-Hispanic White", 46))

  predV = predict(base_logit, newdata=dfimpute)
  # lines(40:85, predV)
  qplot(40:85,predV,geom = "line") + theme_bw()

 ##now look at BMI
  yBMI = 14:80
  dfBMIimpute = data.frame(RIDAGEYR=rep(60,67) , RIAGENDR=rep("Male", 67), BMXBMI=yBMI, DMDEDUC2=rep("HS", 67), RIDRETH1=rep("Non-Hispanic White", 67))
  predBMI = predict(base_logit, newdata=dfBMIimpute)
  qplot(14:80,predBMI,geom = "line") + theme_bw()
```
