---
title: "matrices"
# author: "Laha Ale"
# date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Implement Matrices
#### Categorical Cross Entropy


$$\begin{equation}
L_{\mathrm{CE}}=-\sum_{i=1}^{n} t_{i} \log \left(p_{i}\right), \text { for } \mathrm{n} \text { classes }
\end{equation}$$


Where $t_i$ the true label and $p_i$ is the [softmax](https://en.wikipedia.org/wiki/Softmax_function) probability for the $i^{th}$ class

More details about [Categorical Cross Entropy](https://towardsdatascience.com/cross-entropy-loss-function-f38c4ec8643e)

```{r softmax}
eplison_loss <- 1e-7 # to avoid issue raise by log(0):-Inf
cat_cross_entropy <- function(y_true, y_pred){
  loss <- rowMeans(y_true * log(y_pred + eplison_loss))
  sum(-loss)
}
```

#### Binary Cross-Entropy Loss

$$\begin{equation}
L=-\frac{1}{N}\left[\sum_{j=1}^{N}\left[t_{j} \log \left(p_{j}\right)+\left(1-t_{j}\right) \log \left(1-p_{j}\right)\right]\right]
\end{equation}$$


Where $t_i$ the true label and $p_i$ is the [sigmoid](https://en.wikipedia.org/wiki/Sigmoid_function) probability for the $i^{th}$ class

```{r binary_corss}
bin_cross_entropy <- function(y_true, y_pred){
  loss  <- y_true * log(y_pred)  + (1 - y_true) * log(1-y_pred + eplison_loss)  
  sum(-loss)
}
```

#### Mean squared error

$$\begin{equation}
\mathrm{MSE}=\frac{1}{n} \sum_{i=1}^{n}\left(Y_{i}-\hat{Y}_{i}\right)^{2}
\end{equation}$$


```{r binary_mse}
mean_square_error <- function(y_true, y_pred){
    mean((y_true - y_pred)^2)
}
```

### Load Mice
```{r load_mice, echo = TRUE,message=FALSE}
require(mice)
require(lattice)
library(dplyr)

library(rMIDAS)
# set_python_env(python ="/opt/anaconda3/bin/python")
set_python_env(x ="C:\\ProgramData\\Anaconda3\\",type = "conda")
```

### Load Data

```{r load_data, echo = TRUE}
head(fdgs)
```
#### drop colunms
```{r load_drop_cols, echo = TRUE}
data <- fdgs %>% select(-c(id,wgt.z,hgt.z)) %>% na.omit()
head(data)
```

#### Create Missing Data
```{r create_miss, echo = TRUE}
miss_data <- add_missingness(data, prop = 0.1)
# view miss number of miss data by columns
print(sapply(miss_data, function(x) sum(is.na(x))))
# md.pattern(miss_data)
```

#### Imputing Data with missRanger
```{r imputed_ranger, echo = TRUE}
library(missRanger)
impt_ranger_data <- missRanger(miss_data, num.trees = 100, verbose = 0)
impt_ranger_data$reg <- as.integer(impt_ranger_data$reg) # ranger produce float number of categorical 
impt_ranger_data$sex <- as.integer(impt_ranger_data$sex)

```

#### Imputing Data with MICE
```{r imputed_mice, echo = TRUE}
miss_data <- add_missingness(data, prop = 0.1)
imp <-  mice(miss_data, print=F)
meth <- imp$meth
meth[c('sex','reg')] <- "rf"
meth[c('age','hgt','wgt')] <- 'norm'

imp <- mice(miss_data, m=10, meth = meth, print=F)
imp20  <-  mice.mids(imp, maxit=45, print=F)
plot(imp20)
impt_mice_data <- mice::complete(imp20)
```

#### Imputing Data with RMIDAS
```{r imputed_rmdas, echo = TRUE}
col_bin <- c('sex')
col_cat <- c('reg') 
# Apply rMIDAS preprocessing steps
data_conv <- rMIDAS::convert(miss_data, 
                      bin_cols = col_bin, 
                      cat_cols = col_cat,
                      minmax_scale = TRUE)

# Train the model for 20 epochs
rmidas_train <- rMIDAS::train(data_conv,
                     training_epochs = 50,
                     layer_structure = c(128,128),
                     input_drop = 0.75,
                     seed = 89)

# Generate 10 imputed datasets
impt_rmidas_data <- rMIDAS::complete(rmidas_train, m = 10,fast = TRUE)

```



#### plot categorical loss
```{r plot_categorical, echo = TRUE}

library(ggplot2)

cat_matrix <- function(df,miss_df, impt_df,cols,impt_meth="MICE") 
{
  entory_list <- c()
  
  
  for (col in cols)
  {
    # we only need to compare the missing values
    miss_index <- which(is.na(miss_df[,col]))
    
    
    true_col <- as.factor(df[,col])
    impt_col <- as.factor(impt_df[,col])
    # print(impt_col)
    # NOTE: the imputed data set may have less  levels than the original data
    true_col <-  model.matrix(~0+true_col)
    impt_col <-  model.matrix(~0+impt_col)
    cross_entropy <- cat_cross_entropy(true_col[miss_index,],impt_col[miss_index,])
    entory_list <- c(entory_list,cross_entropy)
    rs <- paste(col,":","categorical crossentropy=",cross_entropy,"method=",impt_meth)
    print(rs)
    
  }
  
  
  matrix_df <- data.frame(categorical=cols, 
                          cross_entropy=cross_entropy,
                          methods=c(rep(impt_meth , length(cols)))
  )
  
  matrix_df
}


comp_cat <- function(data,miss_data,impt_list,cols,methods){
  cat_df <- data.frame(categorical=c(), 
                       cross_entropy=c(),
                       methods=c()
  )
  for (i in 1:length(methods)){
    df <- cat_matrix(data,miss_data,impt_list[[i]],cols,methods[i])
    cat_df <- rbind(cat_df,df)
  }
  
  ggplot(cat_df, aes(fill=methods, y=cross_entropy, x=categorical)) +
    geom_bar(position="dodge", stat="identity")+
    xlab("categorical colunms")+ylab("cross entropy loss")
  
}


```

```{r plot_categorical1, echo = TRUE}

imputed_dataframes <- list(impt_mice_data,impt_rmidas_data[[1]],impt_ranger_data)
miss_data <- as.data.frame(miss_data)
comp_cat(data,miss_data,imputed_dataframes,cols=c("sex","reg"),methods = c("MICE","MIDAS","missRanger"))

```

#### plot numerical loss
```{r plot_mse, echo = TRUE}
num_matrix <- function(df,miss_df, impt_df,cols,impt_meth="MICE") 
{
  entory_list <- c()
  for (col in cols)
  {
    # we only need to compare the missing values
    miss_index <- which(is.na(miss_df[,col]))
    mse <- mean_square_error(df[miss_index,col],impt_df[miss_index,col])
    rs <- paste(col,":","Mean Square Error=",mse,"method=",impt_meth)
    print(rs)
  }
  matrix_df <- data.frame(num=cols, mse=mse, methods=c(rep(impt_meth , length(cols))))
  matrix_df
}


comp_num <- function(data,miss_data,impt_list,cols,methods){
  num_df <- data.frame(num=c(), 
                       mse=c(),
                       methods=c()
  )
  for (i in 1:length(methods)){
    df <- num_matrix(data,miss_data,impt_list[[i]],cols,methods[i])
    num_df <- rbind(num_df,df)
  }
  
  ggplot(num_df, aes(fill=methods, y=mse, x=num)) +
    geom_bar(position="dodge", stat="identity")+
    xlab("Numerical Colunms")+ylab("Mean Square Error")
  
}


```

```{r plot_mse1, echo = TRUE}
comp_num(data,miss_data,imputed_dataframes,cols=c( "age", "hgt","wgt"),methods = c("MICE","MIDAS","missRanger"))

```








