---
title: "More Compare"
author: "Laha Ale"
date: '2022-03-03'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imputation
```{r impt,message=FALSE}
source("imputation.R", local = knitr::knit_global())
```



#### NAs stats in missing dataset

```{r na_stats,message=FALSE}

wgt_nas <- plot_na_pie("wgt")
hgt_nas <- plot_na_pie("hgt")
age_nas <- plot_na_pie("age")
```



## MICE: Wight

#### MICE:compare the imputed datasets with orignal dataset

```{r mice_method,message=FALSE}
df_mice_wgt <- create_compare_data(data,miss_data,impt_mice_data,nas=wgt_nas,
                                   col = "wgt",method = "mice",sp_impt="method")
ggplot(df_mice_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_wgt, aes(source,wgt, colour = source))+geom_boxplot()
ggplot(df_mice_wgt, aes(source,wgt, colour = source))+geom_boxplot(aes(colour=sex))

```


#### MICE:compare split with Sex

```{r mice_sex,message=FALSE}
df_mice_wgt <- create_compare_data(data,miss_data,impt_mice_data,nas=wgt_nas,col = "wgt",method = "mice",sp_impt="sex")
ggplot(df_mice_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_wgt, aes(source,wgt, colour = source))+geom_boxplot()

```

#### MICE:compare by NA counts

```{r mice_NAs_wgt,message=FALSE}


ggplot(df_mice_wgt, aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_wgt, aes(na_count,wgt, colour = sex))+geom_boxplot()


ggplot(df_mice_wgt[grepl("4:|True",df_mice_wgt$na_count),], aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()


```


#### MICE:compare split with age 14

```{r mice_age14,message=FALSE}
# age above 14
ggplot(df_mice_wgt[df_mice_wgt$age >= 14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_mice_wgt[df_mice_wgt$age <14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```


## Ranger: Wight

#### Ranger:compare the imputed datasets with orignal dataset

```{r Ranger_method,message=FALSE}
df_ranger_wgt <- create_compare_data(data,miss_data,impt_ranger_data,nas=wgt_nas,col = "wgt",method = "ranger",sp_impt="method")
ggplot(df_ranger_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_wgt, aes(source,wgt, colour = source))+geom_boxplot()
ggplot(df_ranger_wgt, aes(source,wgt, colour = source))+geom_boxplot(aes(colour=sex))
```


#### Ranger:compare split with Sex

```{r ranger_sex,message=FALSE}
df_ranger_wgt <- create_compare_data(data,miss_data,impt_ranger_data,nas=wgt_nas,col = "wgt",method = "ranger",sp_impt="sex")
ggplot(df_ranger_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_wgt, aes(source,wgt, colour = source))+geom_boxplot()

```


#### Ranger:compare by NA counts

```{r ranger_NAs_wgt}
ggplot(df_ranger_wgt, aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_wgt, aes(na_count,wgt, colour = sex))+geom_boxplot()
ggplot(df_ranger_wgt[grepl("4:|True",df_ranger_wgt$na_count),], aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()


```


#### Ranger:compare split with age 14

```{r ranger_age14}
# age above 14
ggplot(df_ranger_wgt[df_ranger_wgt$age >= 14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_ranger_wgt[df_ranger_wgt$age <14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```


## MIDAS: Wight

#### MIDAS:compare the imputed datasets with orignal dataset

```{r midas_method,message=FALSE}
df_midas_wgt <- create_compare_data(data,miss_data,impt_rmidas_data,nas=wgt_nas,col = "wgt",method = "midas",sp_impt="method")
ggplot(df_midas_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_wgt, aes(source,wgt, colour = source))+geom_boxplot()
ggplot(df_midas_wgt, aes(source,wgt, colour = source))+geom_boxplot(aes(colour=sex))
```


#### MIDAS:compare split with Sex

```{r midas_sex,message=FALSE}
df_midas_wgt <- create_compare_data(data,miss_data,impt_rmidas_data,nas=wgt_nas,col = "wgt",method = "midas",sp_impt="sex")
ggplot(df_midas_wgt, aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_wgt, aes(source,wgt, colour = source))+geom_boxplot()

```




#### MIDAS:compare by NA counts

```{r midas_NAs_wgt,message=FALSE}
ggplot(df_midas_wgt, aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_wgt, aes(na_count,wgt, colour = sex))+geom_boxplot()
ggplot(df_midas_wgt[grepl("4:|True",df_midas_wgt$na_count),], aes(age,wgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()

```


#### MIDAS:compare split with age 14

```{r midas_age14,message=FALSE}
# age above 14
ggplot(df_midas_wgt[df_midas_wgt$age >= 14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_midas_wgt[df_midas_wgt$age <14,], aes(age,wgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```


## MICE: Hight

#### MICE:compare the imputed datasets with orignal dataset

```{r hgt_mice_method,message=FALSE}
df_mice_hgt <- create_compare_data(data,miss_data,impt_mice_data,nas=hgt_nas,col = "hgt",method = "mice",sp_impt="method")
ggplot(df_mice_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_hgt, aes(source,hgt, colour = source))+geom_boxplot()
ggplot(df_mice_hgt, aes(source,hgt, colour = source))+geom_boxplot(aes(colour=sex))
```


#### MICE:compare split with Sex

```{r hgt_mice_sex,message=FALSE}
df_mice_hgt <- create_compare_data(data,miss_data,impt_mice_data,nas=hgt_nas,col = "hgt",method = "mice",sp_impt="sex")
ggplot(df_mice_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_hgt, aes(source,hgt, colour = source))+geom_boxplot()

```


#### MICE:compare by NA counts

```{r mice_NAs_ght,message=FALSE}
ggplot(df_mice_hgt, aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_mice_hgt, aes(na_count,hgt, colour = source))+geom_boxplot()
ggplot(df_mice_hgt[grepl("4:|True",df_mice_hgt$na_count),], aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()


```


#### MICE:compare split with age 14

```{r hgt_mice_age14,message=FALSE}
# age above 14
ggplot(df_mice_hgt[df_mice_hgt$age >= 14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_mice_hgt[df_mice_hgt$age <14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```


## Ranger: Hight

#### Ranger:compare the imputed datasets with orignal dataset

```{r hgt_Ranger_method,message=FALSE}
df_ranger_hgt <- create_compare_data(data,miss_data,impt_ranger_data,nas=hgt_nas,col = "hgt",method = "ranger",sp_impt="method")
ggplot(df_ranger_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_hgt, aes(source,hgt, colour = source))+geom_boxplot()
ggplot(df_ranger_hgt, aes(source,hgt, colour = source))+geom_boxplot(aes(colour=sex))
```


#### Ranger:compare split with Sex

```{r hgt_ranger_sex,message=FALSE}
df_ranger_hgt <- create_compare_data(data,miss_data,impt_ranger_data,nas=hgt_nas,col = "hgt",method = "ranger",sp_impt="sex")
ggplot(df_ranger_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_hgt, aes(source,hgt, colour = source))+geom_boxplot()

```

#### Ranger:compare by NA counts

```{r ranger_NAs_ght,message=FALSE}
ggplot(df_ranger_hgt, aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_ranger_hgt, aes(na_count,hgt, colour = sex))+geom_boxplot()
ggplot(df_ranger_hgt[grepl("4:|True",df_ranger_hgt$na_count),], aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()

```

#### Ranger:compare split with age 14

```{r hgt_ranger_age14,message=FALSE}
# age above 14
ggplot(df_ranger_hgt[df_ranger_hgt$age >= 14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_ranger_hgt[df_ranger_hgt$age <14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```


## MIDAS: Hight

#### MIDAS:compare the imputed datasets with orignal dataset

```{r hgt_midas_method,message=FALSE}
df_midas_hgt <- create_compare_data(data,miss_data,impt_rmidas_data,nas=hgt_nas,col = "hgt",method = "midas",sp_impt="method")
ggplot(df_midas_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_hgt, aes(source,hgt, colour = source))+geom_boxplot()
ggplot(df_midas_hgt, aes(source,hgt, colour = source))+geom_boxplot(aes(colour=sex))
```


#### MIDAS:compare split with Sex

```{r hgt_midas_sex,message=FALSE}
df_midas_hgt <- create_compare_data(data,miss_data,impt_rmidas_data,nas=hgt_nas,col = "hgt",method = "midas",sp_impt="sex")
ggplot(df_midas_hgt, aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_hgt, aes(source,hgt, colour = source))+geom_boxplot()


```

#### MIDAS:compare by NA counts

```{r midas_NAs_ght,message=FALSE}
ggplot(df_midas_hgt, aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()
ggplot(df_midas_hgt, aes(na_count,hgt, colour = sex))+geom_boxplot()


ggplot(df_midas_hgt[grepl("4:|True",df_midas_hgt$na_count),], aes(age,hgt, colour = na_count))+geom_point(alpha=0.4)+stat_smooth()


```

#### MIDAS:compare split with age 14

```{r hgt_midas_age14,message=FALSE}
# age above 14
ggplot(df_midas_hgt[df_midas_hgt$age >= 14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

# age below 14
ggplot(df_midas_hgt[df_midas_hgt$age <14,], aes(age,hgt, colour = source))+geom_point(alpha=0.4)+stat_smooth()

```

# compare miss to true data:wgt

```{r wgt_miss_true_wgt,messwgt=FALSE}
miss_index <- which(is.na(miss_data$wgt))
for (i in 1:10){
  sex <- factor(data$sex[miss_index])
  g1 <- qplot(data$wgt[miss_index],impt_mice_data[[3]]$wgt[miss_index],col=sex)+stat_smooth()+ylim(-10, 105)+
    ylab("mice wgt") + xlab("data wgt")+theme(legend.position = "top")
  
  g2 <- qplot(data$wgt[miss_index],impt_ranger_data[[3]]$wgt[miss_index],col=sex)+stat_smooth()+ylim(-10, 105)+
    ylab("ranger wgt") + xlab("data wgt")+theme(legend.position = "top")
  
  g3 <- qplot(data$wgt[miss_index],impt_rmidas_data[[3]]$wgt[miss_index],col=sex)+stat_smooth()+ylim(-10, 105)+
    ylab("midas wgt") + xlab("data wgt")+theme(legend.position = "top")
  grid.arrange(g1, g2,g3, ncol=3)
  
}
```

# compare miss to true data:hgt

```{r hgt_miss_true_ght,message=FALSE}
miss_index <- which(is.na(miss_data$hgt))
for (i in 1:10){
  sex <- factor(data$sex[miss_index])
  g1 <- qplot(data$hgt[miss_index],impt_mice_data[[3]]$hgt[miss_index],col=sex)+stat_smooth()+ylim(30, 215)+
    ylab("mice hgt") + xlab("data hgt")+theme(legend.position = "top")
  
  g2 <- qplot(data$hgt[miss_index],impt_ranger_data[[3]]$hgt[miss_index],col=sex)+stat_smooth()+ylim(30, 215)+
    ylab("ranger hgt") + xlab("data hgt")+theme(legend.position = "top")
  
  g3 <- qplot(data$hgt[miss_index],impt_rmidas_data[[3]]$hgt[miss_index],col=sex)+stat_smooth()+ylim(30, 215)+
    ylab("midas hgt") + xlab("data hgt")+theme(legend.position = "top")
  grid.arrange(g1, g2,g3, ncol=3)
  
}
```


# compare miss to true data:age

```{r age_miss_true_age,message=FALSE,warning=FALSE}
miss_index <- which(is.na(miss_data$age))
for (i in 1:10){
  sex <- factor(data$sex[miss_index])
  g1 <- qplot(data$age[miss_index],impt_mice_data[[3]]$age[miss_index],col=sex)+stat_smooth()+ylim(-5,22)+
    ylab("mice age") + xlab("data age")+theme(legend.position = "top")
  
  g2 <- qplot(data$age[miss_index],impt_ranger_data[[3]]$age[miss_index],col=sex)+stat_smooth()+ylim(-5,22)+
    ylab("ranger age") + xlab("data age")+theme(legend.position = "top")
  
  g3 <- qplot(data$age[miss_index],impt_rmidas_data[[3]]$age[miss_index],col=sex)+stat_smooth()+ylim(-5,22)+
    ylab("midas age") + xlab("data age")+theme(legend.position = "top")
  grid.arrange(g1, g2,g3, ncol=3)
  
}
```