---
title: "Cobalt Paper"
date: "Updated on : `r date()`"
output: html_document
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Load libs

The goal of this vignette is to replicate the analysis presented in ``Association of blood cobalt concentrations with dyslipidemia, hypertension, and diabetes in a US population
A cross-sectional study; Hongxin Wang, MD, Feng Li, MD, Jianghua Xue, MD, Yanshuang Li, MD, Jiyu Li, MD''

```{r setup,warning=FALSE,message=FALSE}
library(splines)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(nhanesA)
library(phonto)
```


## 2. Data
```{r data_config,warning=FALSE,message=FALSE}
sqlHost <- "localhost"
sqlUserName <- "sa"
sqlPassword <- "yourStrong(!)Password"
sqlDefaultDb <- "NhanesLandingZone"
```
#### 1) Blood Pressure Data
```{r data_blood,warning=FALSE,message=FALSE}
tablnames = searchTableByName('BPX[_]')
tablnames
blood_df <- unionQuery(tablnames$TableName,cols =c("BPXDI1","BPXDI2","BPXSY1","BPXSY2"))

```

#### e) Demo, Body Mesures, and Blood Pressure & Cholesterol 

```{r demo_body,warning=FALSE,message=FALSE}
searchTableByName('BMX')
searchTableByName('DEMO')
# searchTablesByVar('BPQ050A')
searchTablesByVar('BPQ020')
searchTableByName('BPQ')

searchTablesByVar('LBDBCOSI')
searchTablesByVar('LBXBCO')
searchTableByName('CRCO')

# find tables with the above functions and joint query like:
patterns <- c("DemographicVariablesAndSampleWeights","ChromiumAndCobalt","BloodPressureAndCholesterol","BodyMeasures")
cols <- c("RIDAGEYR","RIAGENDR","BMXBMI","RIDRETH1","DMDEDUC2","years",'BPQ050A','BPQ020',"LBDBCOSI","LBXBCO")
base_df <- jointQuery(patterns,cols)
```

```{r data,warning=FALSE,message=FALSE}

data <- merge(blood_df,base_df,by="SEQN")
# it would convert to ordered if we do not make the as factor because all those values are presented by integers in NHANES
data$years <- as.factor(data$years)
##FIXME: use nhanesA here  - we need to have real factor levels not integers
# It is extremely slow in docker and often shows : Timeout was reached: No data pulled,
# I will manually set the factors before we replicate the function
# data = nhanesTranslate('DEMO_D', c('RIAGENDR', 'RIDRETH1'), data=data)
data$RIAGENDR <- as.factor(data$RIAGENDR)
levels(data$RIAGENDR) <- c("Male","Female")
data$RIDRETH1 <- as.factor(data$RIDRETH1)
levels(data$RIDRETH1) <- c('Mexican American','Mexican American','Non-Hispanic White','Non-Hispanic Black','Other Race - Including Multi-Racial')

nDEDUC = ifelse(data$DMDEDUC2 < 3, "<HS", ifelse(data$DMDEDUC2 == 3, "HS", 
                                                    ifelse(data$DMDEDUC2 < 6, ">HS", NA)))
data$DMDEDUC2 <- as.factor(nDEDUC)

 #the variables BPQ050A and BPQ040A are taken conditionally so they have missing values put in
 # for anyone who answered no to BPQ020 - and we need to fix that - since there answer to 50A 
 # had they been asked, would have been no (one presumes)




 #set up a dummy for the ones we want to change
  gg = is.na(data$BPQ050A) & data$BPQ020==2
  bpq50A = data$BPQ050A
  bpq50A[gg] = 2

 ##check it worked
 ##table(bpq50A, data$BPQ050A, useNA="always")

data <- data[data$RIDAGEYR>20,]
data$BPQ050A <- NULL # remove it because we do not use it the fellowing code

data <- na.omit(data)




data <- data[data$BPXDI1!=0 | data$BPXDI2!=0,]
data$DIASTOLIC <- (data$BPXDI1+data$BPXDI2)/2
data$SYSTOLIC <- (data$BPXSY1+data$BPXSY2)/2
```


## 2.1 Hypertension
 We follow the definition they described in the paper, using reported (averaged over 2 measurements) systolic and diastolic blood pressure measurements
as well as self-reported statements regarding whether a physician had ever told them that they have high blood pressure.  One might also look at the use
of prescribed hypertensives, as these will modulate the systolic and diastolic measures.  Data on self-report come from the BPQ tables in NHANES.
https://wwwn.cdc.gov/nchs/nhanes/2011-2012/BPQ_G.htm

```{r Hypertension,warning=FALSE,message=FALSE}
# Hypertension was defined as systolic blood pressure (SBP) ≥140 mm Hg, diastolic blood pressure ≥90mm Hg, or the use of antihypertensive medication. 

  # FIXME: you want to use BPQ040 or BPQ050 I think - are they taking medication, not the report of their physician.  Probably you want 050 - so they are
# taking it now and that would effect the BP as measured.
data$hypertension <- data$DIASTOLIC >= 90 | data$SYSTOLIC >= 140 |  data$BPQ020==1
barplot(table(data$hypertension))
# 
#   FIXME: they report: The current study included 6866 US adults aged 40 years or older. 
#     We need to restrict to age 40 and older.  Do we have the same number of participants?
#   FIXME: could you see how close we are to matching the first few rows in Table 2 of the Cobalt paper?
# 
# Hypertension was defined as systolic blood pressure (SBP) ≥140 mm Hg, diastolic blood pressure ≥90mm Hg, or the use of antihypertensive medication. 
# Diabetes was defined as fasting glucose > 110 mg/dl, hemoglobin A1c (HbA1c) > 6.5% or a history of diabetes. 
# High TC levels were defined as levels >200 mg/dl. 
# High LDL-C levels were defined as levels >130 mg/ dl. 
# Low HDL-C levels were defined as levels <40 mg/dl in males and <50 mg/dl in females. 
# High TG levels were defined as levels >150 mg/dl.

```
## 2.2 Their findings
FIXME: needs some rewriting.
Using the data collected from the National Health and Nutrition Examination Survey (2015-2018), we performed logistic regression to explore the association of blood cobalt 
concentrations with total cholesterol (TC), low-density lipoprotein cholesterol (LDL-C), high- density lipoprotein cholesterol (HDL-C), triglycerides, hypertension, and diabetes.
A total of 6866 adults were included in this study. Participants with higher blood cobalt levels appeared to be older and have a lower body mass index and, were more likely to be 
female (P for trend < .05). After fully adjusting for demographic characteristics (Model 2), compared with the lowest quartile, the highest quartile of blood cobalt 
concentrations had lower odds ratios (ORs) for elevated TC [OR: 0.62, 95% confidential interval (CI): 0.53 to 0.72, P<.001], 
elevated LDL-C (OR: 0.65, 95% CI: 0.53-0.80, P<.001) and low HDL-C (OR: 0.81, 95% CI: 0.69-0.96, P = .013). 
The adjusted ORs for elevated TC, elevated LDL-C and low HDL-C were negatively correlated with increased blood cobalt concentrations (P for trend < .05). 
The adjusted ORs for hypertension and diabetes were not associated with blood cobalt concentrations (P>.05 and P for trend>.05).
In conclusion, higher blood cobalt concentrations were associated with a lower risk of dyslipidemia. 
However, blood cobalt concentrations were not associated with the risk of hypertension or diabetes.

## 3. Model

The paper does not say that they use interaction or transformation, but I tried them, and nothing new has been found in showing the following model comparison.

```{r model1,warning=FALSE,message=FALSE}
base_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+LBDBCOSI, data = data, family = "binomial")
colt_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+LBXBCO, data = data, family = "binomial")


sjPlot::tab_model(base_logit,colt_logit,
                  dv.labels = c("LBDBCOSI", "LBXBCO"),
                  show.ci = FALSE,show.stat = TRUE,show.se = TRUE,p.style = "scientific", digits.p = 2)

```
  FIXME: below is code from RG - we probably want to be using spline functions for AGE and BMI - and then checking those functions.
  We are missing a bunch of the covariates in the Cobalt paper - specifically things like triglycerides etc from Table 2 in that paper.
Those would be interesting to look at, as they seem to have an unusual patter of association.

```{r model2,warning=FALSE,message=FALSE}
   base_logit <- glm(hypertension ~ RIDAGEYR + RIAGENDR + BMXBMI+DMDEDUC2+RIDRETH1 + sqrt(LBXBCO), data = data, family = "binomial")
   base_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7) + RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1 + ns(sqrt(LBXBCO), df=7) + years, data = data, family = "binomial")
   base_logit <- glm(hypertension ~ ns(RIDAGEYR,df=7) + RIAGENDR + ns(BMXBMI,df=7) + DMDEDUC2 + RIDRETH1, data = data, family = "binomial")

   ## try to do some prediction - and then get a plot of the age spline
  ##46 of these
  yvals = seq(40,85,by=1)
  dfimpute = data.frame(RIDAGEYR=yvals, RIAGENDR=rep("Male", 46), BMXBMI=rep(28.9, 46), DMDEDUC2=rep("HS", 46), RIDRETH1=rep("Non-Hispanic White", 46))

  predV = predict(base_logit, newdata=dfimpute)
  # lines(40:85, predV)
  qplot(40:85,predV,geom = "line") + theme_bw()

 ##now look at BMI
  yBMI = 14:80
  dfBMIimpute = data.frame(RIDAGEYR=rep(60,67) , RIAGENDR=rep("Male", 67), BMXBMI=yBMI, DMDEDUC2=rep("HS", 67), RIDRETH1=rep("Non-Hispanic White", 67))
  predBMI = predict(base_logit, newdata=dfBMIimpute)
  qplot(14:80,predBMI,geom = "line") + theme_bw()
```
